---
title: "Data Transformation & Data Visualisation Homework"
author: "Shinin Varongchayakul"
date: "`r Sys.Date()`"
output: pdf_document
---

# This homework is part of Data Science Bootcamp 10 (2024), DataRockie School.

## 1. Instruction
Homework 1
1. Write 5 codes to query data from `nycflights13`
2. Create an R notebook with R Markdown
3. Submit the R notebook in either .html of .pdf format on Discord

Homework 2
1. Write 5 codes to create graphs with the `ggplots` package using any dataset
2. Create an R notebook with R Markdown
3. Submit the R notebook in either .html of .pdf format on Discord

Note:  For Homework 2, I chose to use `nycflights13` so that Homework 1 and 2 could be completed together.


## 2. Overview of "nycflights13"
"nycflights13" is a dataset including all flights departing from airports in New York City (e.g., JFK, LGA) in 2013.


## 3. The Steps
The homework followed these 7 steps.
1. Install and load R packages
2. Preview the dataset
3. Transform and join the datasets
4. Write query #1
5. Write query #2
6. Write query #3
7. Write query #4
8. Write query #5


## 4. The Codes & Findings
### 4.1 Step 1 - Install and Load R Packages
For this homework, I used the following packages:
```{r}
## step 1 - install and load packages
# 1. install packages
install.packages("tidyverse") # for data transformation
install.packages("ggplot2") # for data visualisation
install.packages("lubridate") # to work with date
install.packages("nycflights13") # for the dataset

# 2. load packages
library(tidyverse)
library(ggplot2)
library(lubridate)
library(nycflights13)
```


### 4.2 Step 2 - Preview the Dataset
The following code was used to preview the dataset:
```{r}
## step 2 - preview the datasets
str(flights) # overview of the dataset "flights"
str(airlines) # overview of the dataset "airlines"
str(airports) # overview of the dataset "airports"
?flights # help with the dataset "flights"
?airlines # help with the dataset "airlines"
?airports # help with the dataset "airports"
```


### 4.3 Step 3 - Transform and Join the Datasets
Because the queries required identifying the airlines airports by names, I joined 3 datasets, "flights" and "airlines," and "flights" and "airports" from "nycflights13" to form 3 new datasets which contained the names of the airlines, origin aiports, and destination airports, respectively.
```{r}
## step 3 - transform and join datasets
# 1. create a column "month_name" in "flights"
flights_month <- flights |>
  mutate(month_name = factor(month,
                             levels = 1:12,
                             labels = month.name))


# 2. "flights" vs "airlines"
flights_airlines <- left_join(flights_month,
                              airlines,
                              by = "carrier")

str(flights_airlines) # dataset overview

# 3. "flgihts" vs "airport" for origin
airports_origin <- rename(airports, 
                          origin_name = name)

flights_origin_airports <- left_join(flights_month, 
                                     airports_origin, 
                                     by = c("origin" = "faa"))

str(flights_origin_airports) # dataset overview

# 4. "flgihts" vs "airport" for destination
airports_destination <- rename(airports,
                               destination_name = name)

flights_dest_airports <- left_join(flights_month,
                                   airports_destination,
                                   by = c("dest" = "faa"))

str(flights_dest_airports) # dataset overview
```


### 4.4 Step 4 - Write Query #1
The first query was to find the top 3 airlines with the highest number of flights by month.
#### 4.4.1 Method
```{r}
## step 4 - query 1 - top 3 airlines with the most flights by month
# 1. calculate descriptive statistics for the number of flights
flights_by_airlines <- flights_airlines |>
  select(month_name, carrier, name, flight) |>
  group_by(month_name, carrier, name) |>
  summarise(count_flight = n(), .groups = "drop") |>
  arrange(month_name, desc(count_flight))

flights_by_airlines |> summarise(min = min(count_flight),
                                 max = max(count_flight),
                                 avg = mean(count_flight))
                     
# 2. find the top 3 airlines with most flights in each month
query_1 <- flights_by_airlines |>
  group_by(month_name) |>
  slice_max(count_flight, n = 3)

print(query_1, n = Inf) # print the top 3 airlines

# 3. create a bar graph
query_1 |>
  ggplot(aes(x = reorder(carrier, -count_flight),
             y = count_flight,
             fill = name)) +
  geom_col(position = "dodge") +
  facet_wrap(~ month_name, 
             scales = "free_x") +
  labs(title = "Figure 1. Top 3 Airlines With the Most Flights by Month",
       x = "Airlines (abbr.)",
       y = "Number of Flights",
       fill = "Airlines",
       caption = "Data Package: nycflights13") +
  theme_minimal()
```

#### 4.4.2 Results
1. The total number of flights across the airlines was between 1 and 5,124, with the average of 1,820 flgihts.

2. The top 3 airlines with the highest number of flights in each month are
   (1) United Air Lines (UA)
   (2) JetBlue Airways (B6)
   (3) ExpressJet Airlines (EV)

3. Throughout the year, United Air Lines maintained its number one position.

4. In most of the months, JetBlue Airways came in the second place.

5. However, ExpressJet Airlines took its place in some months such as May and September through November.


### 4.5 Step 5 - Write Query #2
The second query was to find the total number of flights for each origin airport.
#### 4.5.1 Method
```{r}
## step 5 - query 2 - total flights for each origin airport by month
# 1. find the total flights
query_2 <- flights_origin_airports |>
  select(month_name, origin, origin_name, flight) |>
  group_by(month_name, origin, origin_name) |>
  summarise(count_flight = n(), .groups = "drop") |>
  arrange(month_name, origin, origin_name)

print(query_2, n = Inf) # print the total flights

# 2. calculate descriptive statistics
query_2 |>
  summarise(min = min(count_flight),
            max = max(count_flight),
            avg = mean(count_flight))

# 3. create a bar graph
query_2 |>
  ggplot(aes(x = origin,
             y = count_flight,
             fill = origin_name)) +
  geom_col(position = "dodge") +
  facet_wrap(~ month_name, 
             scales = "free_x") +
  labs(title = "The Total Number of Flights for Each Origin Aiports",
       x = "Airports (abbr.)",
       y = "Number of Flights",
       fill = "Airports",
       caption = "Data Package: nycflights13") +
  theme_minimal()
```

##### 4.5.2 Results
1. There were 3 origin airports:
   (1) John F Kennedy Intl (JFK)
   (2) La Guardia (LGA)
   (3) Newark Libery Intl (EWR)
1. The total number of flights at the origin airports was between 7,423 and 10,592 flights, with the average of 9,355 flight.

2. Neward Liberty Intl was ranked first each month in 2013. 

3. La Guardia came in the second from January to August and again in December.

4. John F Kenndy Intl remained in the third place throughout most of the year, except in September, October, and november, where its total flights exceed that at La Guardia.


### 4.6 Step 6 - Write Query #3
The third query was to find the number one destinations with the longest average arrival delay time for each month.
```{r}
## step 6 - query 3 - top 1 destination with the longest average arrival delay by month
# 1. find the top 1 destination with the longest average arrival delay by month
query_3 <- flights_dest_airports |>
  select(month_name, dest, destination_name, arr_delay) |>
  filter(arr_delay > 0) |>
  group_by(month_name, dest, destination_name) |>
  summarise(avg_arr_delay = mean(arr_delay),
            .groups = "drop") |>
  arrange(month_name, desc(avg_arr_delay)) |>
  group_by(month_name) |>
  slice_max(avg_arr_delay, n = 1)

print(query_3, n = Inf) # print the top 1 destination

# 3. create a bar graph
query_3 |>
  ggplot(aes(x = month_name,
             y = avg_arr_delay,
             fill = destination_name)) +
  geom_col(position = "dodge") +
  labs(title = "Destination Airports With the Longest Average Arrival Delay in Each Month",
       x = "Airports (abbr.)",
       y = "Average Arrival Delay (Minutes)",
       fill = "Airports",
       caption = "Data Package: nycflights13") +
  theme_minimal()
```

#### 4.6.2 Results
1. Throughout the year, different destination airports had the longest average arrival delay.

2. The longest delay was in September at Bob Hope (BUR; 122 mintes).

3. The shortest delay was in november at Wilmington Intl (BZN; 52.5 minutes).


### 4.7 Step 7 - Write Query #4
The third query was to find the number one airports with the highest on-time arrival percentage in each month.
#### 4.7.1 Method
```{r}
## step 7 - query 4 - top 1 airport with the highest on-time arrival percentage by month
# 1. find the top 1 airport with the highest on-time arrival percentage by month
query_4 <- flights_dest_airports |>
  select(month_name, dest, destination_name, arr_delay) |>
  mutate(on_time = arr_delay <= 0) |>
  group_by(month_name, dest, destination_name) |>
  summarise(percent_on_time = mean(on_time, na.rm = TRUE)*100,
            .groups = "drop") |>
  arrange(month_name, desc(percent_on_time)) |>
  group_by(month_name) |>
  slice_max(percent_on_time, n = 1)

print(query_4) # print the top 1 airport

# 2. create a bar graph
query_4 |>
  ggplot(aes(x = month_name,
             y = percent_on_time,
             fill = destination_name)) +
  geom_col(position = "dodge") +
  labs(title = "Destination Airports With the Highest On-Time Arrival Percentage in Each Month",
       x = "Airports (abbr.)",
       y = "On-Time Arrival Percentage (%)",
       fill = "Airports",
       caption = "Data Package: nycflights13") +
  theme_minimal()
```

#### 4.7.2 Results
1. When on-time arrival is defined as arrival delay of 0 minutes or less, different airports had the highest on-time arrival each month.

2. The majority of these airports had the on-time arrival ratio of 100%.

3. The lowest percentage was 84.4% at John Wayne Arpt Orange Co in February.

2. Bradley Intl (BDL) was the only airport with the highest on-time arrival in two consecutive months: September (92.3%) and Ocotober (88.9%).

3. No other aiports had attained the first rank more than once.

4. In April, 5 airports were tied for the first place:
   (1) Asheville Regional Airport (AVL; 100%)
   (2) Eagle Co Rgnl (EGE; 100%)
   (3) Key West Intl (EYW; 100%)
   (4) Myrtle Beach Intl (MYR; 100%)
   (5) Palm Springs Intl (PSP; 100%)


### 4.8 Step 8 - Write Query #5
The fifth query was to calculate the total number of flights for each month.
#### 4.8.1 Method
```{r}
## step 8 - query 5 - the number of flights in each month
# 1. calculate the number of flights by month
query_5 <- flights_month |>
  select(month_name, flight) |>
  group_by(month_name) |>
  summarise(count_flight = n())

print(query_5) # print the number of flights in each month

# 2. calculate descriptive statistics
query_5 |> summarise(min = min(count_flight),
                     max = max(count_flight),
                     avg = mean(count_flight))

# 3. create a bar graph
query_5 |>
  ggplot(aes(x = month_name,
             y = count_flight,
             fill = month_name)) +
  geom_col(position = "dodge") +
  labs(title = "The Total Number of Flights (Monthly)",
       x = "Month",
       y = "Flights",
       caption = "Data Package: nycflights13") +
  theme() +
  guides(fill = FALSE)
```

#### 4.8.2 Results
1. The total number of flights in 2013 was between 24,951 and 29,425, with the average of 28,065 flights per month.

2. July had the highest number of flights: 29,425 flights.

3. February had the lowest number of flights: 24,951 flights.

4. Overall, the number of flights from New York City was relatively stable throughout the year.